Internal models of type theory

### 2013-10-29 article-models-tt.tex
universal property for extension
wrapping up the description of type theory

In all of the following inference rules we assume that $\jtype{\ctxext{\Gamma}{A}}{P}$
is among the hypothesis. The induction principle for extension consists of three
inference rules:

\begin{infarray}{c}
\inference{}
{\jtermt{\ctxext{{\Gamma}{A}}{P}}{\ctxwk{P}{{A}{\ctxext{A}{P}}}}{\pair^P}}\\
\inference{
  \jtype{\ctxext{\Gamma}{{A}{P}}}{Q}
  \qquad
  \jtermt{\ctxext{{\Gamma}{A}}{P}}{\subst{\pair^P}{\ctxwk{P}{{A}{Q}}}}{f}}
  {\jtermt{\ctxext{\Gamma}{{A}{P}}}{Q}{\ind{ext}(f)}}\\
\inference{
  \jtype{\ctxext{\Gamma}{{A}{P}}}{Q}
  \qquad
  \jtermt{\ctxext{{\Gamma}{A}}{P}}{\subst{\pair^P}{\ctxwk{P}{{A}{Q}}}}{f}}
  {\jtermteq{\ctxext{{\Gamma}{A}}{P}}{\subst{\pair^P}{\ctxwk{P}{{A}{Q}}}}{\subst{\pair^P}{\ctxwk{P}{{A}{\ind{ext}(f)}}}}{f}}
\end{infarray}

@@

\subsection{The variable rule -- identity functions}
In the usual presentation of type theory, the variable rule asserts that ``in
the context $x_1:A_1,...,x_n:A_n$ we have a term $x_i:A_i$ when $1\leq i\leq n$''.

In our current system, we don't need this generality. Instead, it suffices for
us to assert that there are identity morphisms. We get arbitrary projections
from weakening.

\begin{infarray}{c}
\inference{\jtype{\Gamma}{A}}{\jtermt{\ctxext{\Gamma}{A}}{\ctxwk{A}{A}}{\idfunc[A]}}\\
\inference{\jtermt{\Gamma}{A}{x}}{\jtermteq{\Gamma}{A}{\subst{x}{\idfunc[A]}}{x}}\\
\inference{\jtermt{\ctxext{\Gamma}{A}}{P}{f}}{\jtermteq{\ctxext{\Gamma}{A}}{P}{\subst{\idfunc[A]}{\ctxwk{A}{f}}}{f}}\\
\inference{\jtermt{\ctxext{\Gamma}{B}}{\ctxwk{B}{A}}{g}}{\jtermteq{\ctxext{\Gamma}{B}}{\ctxwk{B}{A}}{\subst{g}{\ctxwk{B}{\idfunc[A]}}}{g}}
\end{infarray}

@@

%\begin{infarray}{c}
%\inference{\jtype{\ctxext{{\Gamma}{A}}{P}}{Q}\qquad\jtermt{\Gamma}{A}{x}\qquad\jtermt{\ctxext{A
%\end{infarray}{c}

\subsection{Compatibility of extension, weakening and substitution}
We describe the compatibility of extension, weakening and substitution. These
compatibility rules are formulated in a weak sense.

Extension is compatible with extension:
\begin{infarray}{c}
\inference
{ \jtype{\ctxext{\Gamma}{A}}{P}
}{\jtermt
  {\ctxext{\Gamma}{{A}{P}}}
  {\ctxwk{\ctxext{\Gamma}{{A}{P}}}{\ctxext{{\Gamma}{A}}{P}}}{\typefont{extext}_0(A,P)}
}\\
\inference
{ \jtype{\ctxext{{\Gamma}{A}}{P}}{Q}
}{\jtermt
  {\ctxext{\Gamma}{{A}{{P}{Q}}}}
  {\ctxwk{\ctxext{A}{{P}{Q}}}{\ctxext{{A}{P}}{Q}}}
  {\typefont{extext}_1(A,P,Q)}
}
\end{infarray}

Weakening is compatible with extension:
\begin{infarray}{c}
\inference{
  \jctx{\Gamma}
  \qquad
  \jtype{\Delta}{B}}
  {\jtermt
    {\ctxext{\Gamma}{\ctxwk{\Gamma}{\ctxext{\Delta}{B}}}}
    {\ctxwk{{\Gamma}{\ctxext{\Delta}{B}}}{\ctxext{\ctxwk{\Gamma}{\Delta}}{\ctxwk{\Gamma}{B}}}}
    {\typefont{wkext}_0(\Gamma,B)}}
\end{infarray}

Substitution is compatible with substitution:
\begin{infarray}{c}
\inference{\jtype{\ctxext{{\Gamma}{A}}{P}}{Q}\qquad\jtermt{\Gamma}{A}{x}\qquad\jtermt{\ctxext{\Gamma}{A}}{P}{f}}
{\jtypeeq{\Gamma}{\subst{x}{\subst{f}{Q}}}{\subst{{x}{f}}{\subst{x}{Q}}}}
\end{infarray}

\subsection{More stuff in the type theory}
\begin{infarray}{c}
\inference{\jtype{\Gamma}{A}}{\jtermt{\ctxext{\Gamma}{A}}{\ctxwk{A}{A}}{\idfunc[A]}}
\end{infarray}

@@

\subsection{The interactions of extension, weakening and substitution}

\subsection{The basic constructors of type theory}
\subsubsection{Dependent function types}
\begin{infarray}{cc}
\inference{\jtype{\ctxext{\Gamma}{A}}{P}}{\jtype{\Gamma}{\mprd{A}{P}}}
& \inference{\jtypeeq{\ctxext{\Gamma}{A}}{P}{P'}}{\jtypeeq{\Gamma}{\mprd{A}{P}}{\mprd{A}{P'}}}\\
\inference{\jtermt{\ctxext{\Gamma}{A}}{P}{f}}{\jtermt{\Gamma}{\mprd{A}{P}}{\lambda(f)}}
& \inference{\jtermteq{\ctxext{\Gamma}{A}}{P}{f}{f'}}{\jtermteq{\Gamma}{\mprd{A}{P}}{\lambda(f)}{\lambda(f')}}\\
\inference{\jtermt{\Gamma}{\mprd{A}{P}}{g}}{\jtermt{\ctxext{\Gamma}{A}}{P}{\tfev(g)}}
& \inference{\jtermteq{\Gamma}{\mprd{A}{P}}{g}{g'}}{\jtermteq{\ctxext{\Gamma}{A}}{P}{\tfev(g)}{\tfev(g')}}\\
\inference{\jtermt{\ctxext{\Gamma}{A}}{P}{f}}{\jtermteq{\Gamma}{A}{\tfev(\lambda(f))}{f}}
\end{infarray}

>>>>

@@

We begin with stating the universal property of the extension $\ctxext{\Gamma}{A}$.
In these rules we assume we have $\jtype{\Gamma}{A}$ in the hypotheses.

\begin{infarray}{c}
\inference{}
{\jtermt{\ctxext{\Gamma}{A}}{\ctxwk{A}{{\Gamma}{\ctxext{\Gamma}{A}}}}{\pair_A}}\\
\inference{
  \jtype{\ctxext{\Gamma}{A}}{P}
  \qquad
  \jtermt{\ctxext{\Gamma}{A}}{\subst{\pair_A}{\ctxwk{A}{{\Gamma}{P}}}}{f}}
  {\jtermt{\ctxext{\Gamma}{A}}{P}{\ind{\tfext_\Gamma(A)}(f)}}\\
\inference{
  \jtype{\ctxext{\Gamma}{A}}{P}
  \qquad
  \jtermt{\ctxext{\Gamma}{A}}{\subst{\pair_A}{\ctxwk{A}{{\Gamma}{P}}}}{f}}
  {\jtermteq{\ctxext{\Gamma}{A}}{\subst{\pair_A}{\ctxwk{A}{{\Gamma}{P}}}}{\subst{\pair_A}{\ctxwk{A}{{\Gamma}{\ind{\tfext_\Gamma(A)}(f)}}}}{f}}
\end{infarray}

Under the hypothesis that $\jtypeeq{\Gamma}{A}{A'}$
we will also have the rules

\begin{infarray}{c}
\inference{}{\jtermteq{\ctxext{\Gamma}{A}}{\ctxwk{A}{{\Gamma}{\ctxext{\Gamma}{A}}}}{\pair_A}{\pair_{A'}}}\\
\inference{\jtype{\ctxext{\Gamma}{A}}{P}\qquad\jtermt{\ctxext{\Gamma}{A}}{\subst{\pair_A}{\ctxwk{A}{{\Gamma}{P}}}}{f}}
{\jtermteq{\ctxext{\Gamma}{A}}{P}{\ind{\tfext_\Gamma(A)}(f)}{\ind{\tfext_\Gamma(A')}(f)}}
\end{infarray}

Note that we don't need the notion of terms for contexts to state the universal
property of context extension (which is a good thing, for we don't assume to have them).

Next, we give the universal property of the extension $\ctxext{A}{P}$ in context
$\Gamma$.
In all of the following inference rules we assume that $\jtype{\ctxext{\Gamma}{A}}{P}$
is among the hypothesis. The induction principle for extension consists of three
inference rules:

\begin{infarray}{c}
\inference{}
{\jtermt{\ctxext{{\Gamma}{A}}{P}}{\ctxwk{P}{{A}{\ctxext{A}{P}}}}{\pair_P}}\\
\inference{
  \jtype{\ctxext{\Gamma}{{A}{P}}}{Q}
  \qquad
  \jtermt{\ctxext{{\Gamma}{A}}{P}}{\subst{\pair_P}{\ctxwk{P}{{A}{Q}}}}{f}}
  {\jtermt{\ctxext{\Gamma}{{A}{P}}}{Q}{\ind{\tfext_A(P)}(f)}}\\
\inference{
  \jtype{\ctxext{\Gamma}{{A}{P}}}{Q}
  \qquad
  \jtermt{\ctxext{{\Gamma}{A}}{P}}{\subst{\pair_P}{\ctxwk{P}{{A}{Q}}}}{f}}
  {\jtermteq{\ctxext{{\Gamma}{A}}{P}}{\subst{\pair_P}{\ctxwk{P}{{A}{Q}}}}{\subst{\pair_P}{\ctxwk{P}{{A}{\ind{\tfext_A(P)}(f)}}}}{f}}
\end{infarray}

As with context extension, we shall require two more inference rules stating that
$\pair_P$ and $\ind{\tfext_A(P)}$ are invariant under judgmental equality.

\subsection{The variable rule -- identity functions}
The variable rule for contexts:
\begin{infarray}{c}
\inference{\jctx{\Gamma}}{\jtermt{\Gamma}{\ctxwk{\Gamma}{\Gamma}}{\idfunc[\Gamma]}}\\
\inference{\jtermt{\Gamma}{A}{x}}{\jtermteq{\Gamma}{A}{\subst{\idfunc[\Gamma]}{\ctxwk{\Gamma}{x}}}{x}}\\
\inference{\jtermt{\Delta}{\ctxwk{\Delta}{\Gamma}}{g}}{\jtermteq{\Delta}{\ctxwk{\Delta}{\Gamma}}{\subst{g}{\ctxwk{\Delta}{\idfunc[\Gamma]}}}{g}}
\end{infarray}

The variable rule for types:
\begin{infarray}{c}
\inference{\jtype{\Gamma}{A}}{\jtermt{\ctxext{\Gamma}{A}}{\ctxwk{A}{A}}{\idfunc[A]}}\\
\inference{\jtermt{\Gamma}{A}{x}}{\jtermteq{\Gamma}{A}{\subst{x}{\idfunc[A]}}{x}}\\
\inference{\jtermt{\ctxext{\Gamma}{A}}{P}{f}}{\jtermteq{\ctxext{\Gamma}{A}}{P}{\subst{\idfunc[A]}{\ctxwk{A}{f}}}{f}}\\
\inference{\jtermt{\ctxext{\Gamma}{B}}{\ctxwk{B}{A}}{g}}{\jtermteq{\ctxext{\Gamma}{B}}{\ctxwk{B}{A}}{\subst{g}{\ctxwk{B}{\idfunc[A]}}}{g}}
\end{infarray}

@@

\subsection{Other type constructors}
We will now describe the rules for the type constructors that we don't assume
in our type theory by default. The first of these are the dependent function
types. We also discuss $\tfW$-types, suspensions, truncations. More general higher inductive
types will have to wait until we have introduced models, for the models of the
basic type theory because we will use them as index categories of the diagrams.

\subsubsection{Dependent function types}
\begin{infarray}{cc}
\inference{\jtype{\ctxext{\Gamma}{A}}{P}}{\jtype{\Gamma}{\mprd{A}{P}}}
& \inference{\jtypeeq{\ctxext{\Gamma}{A}}{P}{P'}}{\jtypeeq{\Gamma}{\mprd{A}{P}}{\mprd{A}{P'}}}\\
\inference{\jtermt{\ctxext{\Gamma}{A}}{P}{f}}{\jtermt{\Gamma}{\mprd{A}{P}}{\lambda(f)}}
& \inference{\jtermteq{\ctxext{\Gamma}{A}}{P}{f}{f'}}{\jtermteq{\Gamma}{\mprd{A}{P}}{\lambda(f)}{\lambda(f')}}\\
\inference{\jtermt{\Gamma}{\mprd{A}{P}}{g}}{\jtermt{\ctxext{\Gamma}{A}}{P}{\tfev(g)}}
& \inference{\jtermteq{\Gamma}{\mprd{A}{P}}{g}{g'}}{\jtermteq{\ctxext{\Gamma}{A}}{P}{\tfev(g)}{\tfev(g')}}\\
\inference{\jtermt{\ctxext{\Gamma}{A}}{P}{f}}{\jtermteq{\Gamma}{A}{\tfev(\lambda(f))}{f}}
\end{infarray}

@@
